---
title: "Survival Analysis HW Chapter 2"
author: "Victor Feagins"
date: "9/8/2021"
always_allow_html: true
output:
  html_document: default
  word_document: default
---

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
library(readxl)
library(dplyr)
library(flextable)
library(ggplot2)
theme_set(theme_bw())
```

## Exercise 2.1

Consider the survival data given in Exercise Table 2.1 and compute and plot the **estimated survivorship**, the **probability density** and the **hazard** functions.

```{r, echo=FALSE}
data.2.1 <- read_excel("Table 2.1 - Page 13 HW.xlsx", sheet = "Exercise 2.1", skip = 1, col_names = c("Initial_Time", "End_Time", "Number_Survivors_Initial", "Number_Dying_in_Interval"))
data.2.1 %>% 
  flextable() %>% 
  autofit()
```

### Survivorship Function
Survivorship is the probability that an individual survives longer than t. It is estimated with the following equation:

$$\hat{S}(t) = \frac{number\;of\;items\;surviving\;longer\;than\;t}{total\;number\;of\;patients}$$

#### Calcuating Survivorship Exercise 2.1

```{r}
data.2.1 <- data.2.1 %>% 
  mutate('S(t)' = Number_Survivors_Initial/max(Number_Survivors_Initial))
data.2.1 %>% 
  select(Initial_Time, "S(t)") %>% 
  flextable() %>% 
  autofit()
```
#### Plotting Survivorship Exercise 2.1
```{r}
ggplot(data.2.1, aes(`Initial_Time`, `S(t)`)) +
  geom_line() +
  geom_point() +
  ylab(expression(paste(hat("S"), "(t)"))) +
  xlab("t")
```

### Probability Density

#### Calcuating probability density  Exercise 2.1
To estimate the probability density the following equation is used:

$$\hat{f}(t) = \frac{number\;of\;items\;dying\;in\;the\;interval\;beginning\;at\; time \;t}{(total\;number\;of\;items)*(interval\; width)}$$
This probability density is at the mid point of the interval
```{r}
data.2.1 <- data.2.1 %>% 
  mutate(Mid_Time = (Initial_Time + End_Time)/2) %>% 
  mutate("f(t)" = Number_Dying_in_Interval/(max(Number_Survivors_Initial) * (End_Time - Initial_Time)))
data.2.1 %>% 
  select(Mid_Time, "f(t)") %>% 
  flextable() %>% 
  autofit()
```

#### Plotting probability density  Exercise 2.1

```{r}
ggplot(data.2.1, aes(`Mid_Time`, `f(t)`)) +
  geom_line() +
  geom_point() +
  ylab(expression(paste(hat("f"), "(t)"))) +
  xlab("t")
```

### Hazard Function
#### Calcuating Hazard Function Exercise 2.1
To estimate the hazard function the following equation can be used:

$$\hat{h}(t) = \frac{number\;of\;items\;dying\;per\;unit\;time\;in\;the\; interval}{(number\;of\;items\;surviving \; at\; t)-(number\;of\;deaths\;in\;the\;interval)/2}$$

```{r}
data.2.1 <- data.2.1 %>% 
  mutate(Deaths_Per_Time = Number_Dying_in_Interval/(End_Time - Initial_Time),
         "h(t)" = Deaths_Per_Time/(Number_Survivors_Initial - Number_Dying_in_Interval/2))

data.2.1 %>% 
  select(Mid_Time, "h(t)") %>% 
  flextable() %>% 
  autofit()
```

#### Plotting Hazard Function Exercise 2.1


```{r}
ggplot(data.2.1, aes(`Mid_Time`, `h(t)`)) +
  geom_line() +
  geom_point() +
  ylab(expression(paste(hat("h"), "(t)"))) +
  xlab("t")
```

## Exercise 2.2
Table 2.2 is a life table for the total population of 100,000 live births in the United States, 1959 – 1961. Compute and plot the estimated **survivorship**, the **probability density**, and the **hazard** function.

```{r, echo=FALSE}
data.2.2 <- read_excel("Table 2.1 - Page 13 HW.xlsx", sheet = "Exercise 2.2", skip = 1, col_names = c("Initial_Time", "End_Time", "Number_Survivors_Initial", "Number_Dying_in_Interval"))
data.2.2 %>% 
  flextable() %>% 
  autofit()
```

### Survivorship Function
#### Calcuating Survivorship Exercise 2.2
```{r}
data.2.2 <- data.2.2 %>% 
  mutate('S(t)' = Number_Survivors_Initial/max(Number_Survivors_Initial))
data.2.2 %>% 
  select(Initial_Time, "S(t)") %>% 
  flextable() %>% 
  autofit()
```

#### Ploting Survivorship Exercise 2.2
```{r}
ggplot(data.2.2, aes(`Initial_Time`, `S(t)`)) +
  geom_line() +
  geom_point() +
  ylab(expression(paste(hat("S"), "(t)"))) +
  xlab("t")
```

### Probability Density
#### Calcuating probability density  Exercise 2.2

```{r}
data.2.2 <- data.2.2 %>% 
  mutate(Mid_Time = (Initial_Time + End_Time)/2) %>% 
  mutate("f(t)" = Number_Dying_in_Interval/(max(Number_Survivors_Initial) * (End_Time - Initial_Time)))
data.2.2 %>% 
  select(Mid_Time, "f(t)") %>% 
  flextable() %>% 
  autofit()
```

#### Plotting probability density  Exercise 2.2

```{r}
ggplot(data.2.2, aes(`Mid_Time`, `f(t)`)) +
  geom_line() +
  geom_point() +
  ylab(expression(paste(hat("f"), "(t)"))) +
  xlab("t")
```

### Hazard Function
#### Calcuating Hazard Function Exercise 2.2
```{r}
data.2.2 <- data.2.2 %>% 
  mutate(Deaths_Per_Time = Number_Dying_in_Interval/(End_Time - Initial_Time),
         "h(t)" = Deaths_Per_Time/(Number_Survivors_Initial - Number_Dying_in_Interval/2))

data.2.2 %>% 
  select(Mid_Time, "h(t)") %>% 
  flextable() %>% 
  autofit()
```

#### Plotting Hazard Function Exercise 2.2
```{r}
ggplot(data.2.2, aes(`Mid_Time`, `h(t)`)) +
  geom_line() +
  geom_point() +
  ylab(expression(paste(hat("h"), "(t)"))) +
  xlab("t")
```

## Exercise 2.5
Given the survivorship function:
$$S(t) = exp(-t^\gamma) $$
derive the probability density function and the hazard function.

### Relationships
We know that the cumulative Hazard function is the negative natural log of the survival function

$$ H(t) =  -log(S(t))$$

We know that the cumulative Hazard function is the integral of the hazard function

$$ H(t) = \int_{0}^{t} h(x) dx $$
$$H^\prime(t) = h(t)$$
We also know that the probability density function can be derived as such

$$f(t) = S(t) * h(t) $$

### Deriving the cumulative hazard function


$$ H(t) = -log(S(t))$$
$$ H(t) = -log(exp(-t^\gamma))$$
$$ H(t) = -(-t^\gamma) $$
$$ H(t) = t^\gamma $$

### Deriving the hazard function


$$ H(t) = t^\gamma $$
$$ \frac{dH(t)}{dt} = \frac{d}{dt}t^\gamma $$
$$ H^\prime(t) = \gamma t^{\gamma -1} $$
$$ h(t) = \gamma t^{\gamma -1} $$


### Deriving the probability density function

$$ f(t) = S(t) * h(t) $$
$$ f(t) = exp(-t^\gamma) * \gamma t^{\gamma -1} $$
$$ f(t) = \gamma * t^{\gamma - 1} * exp(-t^\gamma) $$

