---
title: "Survival Analysis HW 3"
author: "Victor Feagins"
date: "10/1/2021"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
library(dplyr) 
library(survival)
library(flextable)
library(ggplot2)
library(ggfortify)
theme_set(theme_bw()) #Make ggplots black and white theme
```

### Question 5.8

Using the data from Table 3.1 compare the age distribution of the two groups


```{r, echo=FALSE}
df.5.8 <- haven::read_sas("total.sas7bdat")

df.5.8 %>% 
  select(-CENSOR1, - CENSOR2) %>% 
  head(10) %>% 
  flextable() %>% 
  set_caption("First 10 Observations of Melanoma Patients") %>% 
  autofit()
```

#### Lifetime
In this problem we will be looking at lifetime of a patient comparing treatments.
```{r}
surv.data.5.8 <- Surv(df.5.8$LIFETIME, df.5.8$DEATH)
km.5.8 <- survfit(surv.data.5.8 ~ TREATMENT, data = df.5.8)
summary(km.5.8)
```
```{r}
autoplot(km.5.8, conf.int = F) +
  labs(title = "Life Time Survival by Treatment",
       x = "Time",
       y = "Survival Probability",
       colour = "Treatment")
```

Testing to see if the groups are different?

```{r}
survdiff(surv.data.5.8 ~ TREATMENT, data = df.5.8)
```

When looking at the log rank test we have a p value of .4 which means we fail to reject the null that the two treatments result in different lifetime survival times.


#### Remission Time

We did not see a significant change in the lifetimes between treatments but perhaps we can see a difference in the remission times. 

```{r}
surv.data.5.8 <- Surv(df.5.8$REMTIME, df.5.8$RELAPSE)
km.5.8 <- survfit(surv.data.5.8 ~ TREATMENT, data = df.5.8)
summary(km.5.8)
```

```{r}
autoplot(km.5.8, conf.int = F) +
  labs(title = "Time to Relapse by Treatment",
       x = "Time",
       y = "Survival Probability",
       colour = "Treatment")
```

```{r}
survdiff(surv.data.5.8 ~ TREATMENT, data = df.5.8)
```

In time to relapse we also don't see a significant difference in time to relapse with a p value of .8.

### Question 5.14
Sixty-six patients are separated into three treatment groups. The initial remission times are given in weeks. Are the three treatment groups equally effective in prolonging remission? Discuss your findings!

```{r, echo = FALSE}
df.5.14 <- haven::read_sas("example5_14.sas7bdat")

df.5.14 %>% 
  group_by(treat) %>% 
  slice_sample(n = 3) %>% 
  flextable() %>% 
  set_caption("9 Observations of Remission Times")
```

```{r}
surv.data.5.14 <- Surv(df.5.14$time, abs(df.5.14$cens))
km.5.14 <- survfit(surv.data.5.14 ~ treat, data = df.5.14)
summary(km.5.14)
```

```{r}
autoplot(km.5.14, conf.int = F) +
  labs(title = "Remission Times by Treatment",
       x = "Time",
       y = "Survival Probability",
       colour = "Treatment")
```

```{r}
survdiff(surv.data.5.14 ~ treat, data = df.5.14)
```

Over all treatments we see a significance difference in at least one treatment.
```{r}
survminer::pairwise_survdiff(Surv(time, cens) ~ treat, data = df.5.14, p.adjust.method = "bonferroni")
```

When doing pairwise comparisons we see that treatment 1 is the most different which we could have inferred from the graph but none of the treatments are significantly different from each other. We will need more information in treatment 1 study to possibly arrive at statistical significance. 

```{r}
survminer::pairwise_survdiff(Surv(time, cens) ~ treat, data = df.5.14, p.adjust.method = "none")
```

It is important to recognize that without adjusting for the p.value we do see a significant difference showing the importance for adjusting and possible reasoning to continue examining treatment 1. 

