---
title: "Probability with Data"
author: "Victor Feagins"
date: "11/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(MASS)
library(dplyr)
library(ggplot2)
```
```{r}
df <- iris
```

We can fit various distributions to data. If we look at the Sepal.Width in the iris dataset it seems to be normal.

```{r}
ggplot(df, mapping = aes(x = Sepal.Width))+
  geom_histogram(aes(y = ..density..), color="black", fill="white")+
  geom_density()
```

We can do a statistical test to see if it is normal

```{r}
shapiro.test(df$Sepal.Width)
```
We fail to reject the null that the data is not normal so we can assume that it is normal.

Let's fit the normal distributions parameters from the data.
```{r}
norm.parm <- fitdistr(df$Sepal.Width, "normal") #From the MASS package
```


We can now use these parameters in the normal cumulative density function (CDF). Remember the CDF is the probability of a value being less then $x$

$$ F(x) = P(X \le x) $$


For the normal distribution we would expect that 50% of the data is less then the mean of the data.
```{r}
pnorm(mean(df$Sepal.Width), mean = norm.parm$estimate["mean"], sd = norm.parm$estimate["sd"])
```

Using the `qnorm` function we see that the mean lies at the 50% probability. We would expect a 50% chance to see points above or below this point. We can also do $ 1 - P(X \le x) $ to see what the probability of seeing a point greater then x.


