---
title: "Survival Analysis HW 2"
author: "Victor Feagins"
date: "9/20/2021"
output:
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE)
library(readxl)
library(dplyr)
library(flextable)
library(ggplot2)
library(survival)
library(survminer)
library(broom)
library(magrittr)
theme_set(theme_bw())
```

## 4.1 Consider the survival time of the 30 melanoma patients in Table 3.1.
```{r}
df.3.1 <-  read_excel("Table 3.1.xlsx")
head(df.3.1, 10) %>% 
  flextable() %>% 
  set_caption("First 10 Observations")
```

### Compute and plot the PL estimates of the survivorship functions $S(t)$ of the two treatment groups and check your results with Table 3.2 and Figure 3.1.

```{r}
surv.model <- Surv(df.3.1$`Survival Time`, event = df.3.1$Uncensor...12)
```


```{r}
km <- survfit(surv.model ~ Treatment, data = df.3.1)
summary(km)
```
 The survival times are the same with Table 3.2 in the book.

```{r}
ggsurvplot(km)
```

The survival curves are the same as figure 3.1 in the book.


### Compute the variance of $S(t)$ for every uncensored observation.
The following equation is used to calculate the variance of $S(t)$.

$$
Var[\hat{S}(t)] = [\hat{S}(t)]^2 \sum_{r} \frac{1}{(n-r)(n-r + 1)}
$$

Where $n$ is number of observations and $r$ is number of deaths.
```{r}
surv.data <- tidy(km)

surv.var <- surv.data %>%
  filter(n.event == 1) %>% #Looking only at uncensored observations
  group_by(strata) %>% 
  mutate(S.Var = estimate^2 * cumsum(1/((max(n.risk)-cumsum(n.event))*(max(n.risk)-cumsum(n.event) + 1))))

surv.var %>% 
  select(time, strata, S.Var) %>% 
  flextable() %>% 
  set_caption("Surival Function Variences")
```
```{r, include=FALSE}
surv.data <- tidy(km)

test <- surv.data %>%
  filter(n.event == 1) %>% #Looking only at uncensored observations
  group_by(strata) %>% 
  mutate(S2 = estimate^2,
         c.r = cumsum(n.event),
         n = max(n.risk),
         n_r = 1/((n-c.r)*(n-c.r+1)),
         c.n_r = cumsum(n_r),
         var = S2 * c.n_r)

```



### Estimate the median survival times of the two groups

```{r}
print(km)
```
For treatment 0 the median is undefined since the is still more then 50% of patients alive. 

## 4.4 Consider the remission data of 42 patients with acute leukemia in Example 3.3.

```{r, include= FALSE}
MP.data <- c(6,6,6,7,10,13,16,22,23,6,9,10,11,17, 19,20,25,32,32,34,35)
censor.MP <- c(rep(1,9),rep(0,12))
treatment.MP <- rep("6-MP", length(MP.data))
P.data <- c(1,1,2,2,3,4,4,5,5,8,8,8,8,11,11,12,12,15,17,22,23)
censor.P <- rep(1,21)
treatment.P <- rep("Placebo", length(P.data))

df.3.3 <- data.frame(Time = c(P.data, MP.data), Censor= c(censor.P, censor.MP), treatment =c(treatment.P, treatment.MP))

```
```{r, echo=FALSE}
print("Placebo")
df.3.3 %>% filter(treatment == "Placebo") %$%
  Surv(Time, event = Censor)
print("6-MP")
df.3.3 %>% filter(treatment == "6-MP") %$%
  Surv(Time, event = Censor)
```


```{r, include = FALSE}
surv.model.3.3 <- Surv(df.3.3$Time, event = df.3.3$Censor)
```


### Compute and plot the PL estimates of $S(t)$ at every time to relapse for the 6-MP and placebo groups.
```{r}
km <- survfit(surv.model.3.3 ~ treatment, data = df.3.3)
summary(km)
```

```{r}
ggsurvplot(km)
```


### Compute the variances of $S(10)$ in the 6-MP group and of $S(3)$ in the placebo group.

```{r}
surv.data <- tidy(km)

surv.var <- surv.data %>%
  filter(n.event == 1) %>% #Looking only at uncensored observations
  group_by(strata) %>% 
  mutate(S.Var = estimate^2 * cumsum(1/((max(n.risk)-cumsum(n.event))*(max(n.risk)-cumsum(n.event) + 1))))

surv.var %>% 
  select(time, strata, S.Var) %>% 
  filter(time == 10 | time == 3) %>% 
  flextable() %>% 
  set_caption("Surival Function Variences S(10) & S(3)")
```


### Estimate the median remission times of the two treatment groups.
```{r}
print(km)
```

## 4.9 Do a complete life-table analysis of the data given in Exercise Table 4.2, the Angina Pectoris data.
```{r, eval = FALSE, include= FALSE}
df.4.9 <- haven::read_sas("males.sas7bdat")
head(df.4.9) %>% 
  flextable() %>% 
  set_caption("First 5 Observations")
```
```{r}
df.4.9 <- read_excel("Female4.2.xlsx")
head(df.4.9) %>% 
  flextable() %>% 
  set_caption("First 5 Observations")
```


This data is different format of survival data. R handles survival data where each row is an event. So we will need to transform the data by replicated the number of rows by the associated rows frequency 

```{r, eval= FALSE, include= FALSE}
duptimes <- df.4.9$Freq
idx <- rep(1:nrow(df.4.9), duptimes)
df.4.9.t <- df.4.9[idx,]
```
```{r, include=FALSE}
duptimes.d <- df.4.9$Deaths
idd <- rep(1:nrow(df.4.9), duptimes.d)

Times.d <- df.4.9[idd,"Interval"]
data.d <- data.frame(Times = Times.d, Deaths = 1)

duptimes.c <- df.4.9$Censored
idc <-  rep(1:nrow(df.4.9), duptimes.c)

Times.c <- df.4.9[idc,"Interval"]

data.c <- data.frame(Times = Times.c, Deaths = 0)
```


```{r}
df.4.9.t <- rbind(data.d, data.c)
df.4.9.t <- df.4.9.t %>% 
  tidyr::separate(Interval, c("Start", "End"), "—") %>%
  mutate(Deaths = ifelse(is.na(End), 0, Deaths),# if we don't know when they died make it censored
         End = ifelse(is.na(End), 15, End)) 
```
Once that transformation is done survival analysis is the same steps as the previous problems. 
```{r}
surv.model.4.9 <- Surv(#time = as.numeric(df.4.9.t$Start),
                       time = as.numeric(df.4.9.t$End),
                       event = df.4.9.t$Deaths)
```
```{r}
km <- survfit(surv.model.4.9 ~ 1, data = df.4.9.t)
summary(km)
```



### Plot the three survival functions: survival, hazard and density functions.

#### Survival Function

```{r}
ggsurvplot(km)
```

#### Hazard Function
```{r}
survival.table1 <- broom::tidy(km) %>% filter(n.event > 0)
survival.table1 <- survival.table1 %>% mutate(hazard = n.event / (n.risk * (lead(time) - time)))
ggplot() +
  geom_line(data = survival.table1, aes(x = time, y = hazard)) +
  labs(x = "Time", y = "Hazard")
```

### Probablity 
```{r}
survival.table1 <- survival.table1 %>% mutate(probability  = estimate * hazard)
ggplot() +
  geom_line(data = survival.table1, aes(x = time, y = probability)) +
  labs(x = "Time", y = "Probability Density")
```




